---
title: "Experiment"
bibliography: references.bib
link-citations: true
date: "Last compiled on `r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    self_contained: true
    css: tweaks.css
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 4
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test3"))
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

---

# Getting started

To copy the code, click the button in the upper right corner of the code-chunks.

## clean up

```{r, clean_up, results='hide'}
rm(list=ls())
gc()
```

<br>

## custom functions

We defined a number custom functions, at `r xfun::embed_file("./custom_functions.R")`.

```{r, custom_functions}
source("./custom_functions.R")
```

<br>

## necessary packages

- `tidyverse`: data wrangling
- `igraph`: generate and visualize graphs
- `parallel`: parallel computing to speed up simulation
- `foreach`: looping in parallel
- `doParallel`: parallel backend for `foreach`
- `ggplot2`: data visualization
- `ggh4x`: hacks for `ggplot2`
- `ggpubr`: make visualizations publication-ready


```{r, packages}
packages = c("tidyverse", "igraph", "ggplot2", "parallel", "doParallel", "foreach", "ggh4x", "ggpubr", "plotly", "RColorBrewer", "grid", "gridExtra", "patchwork", "ggplotify", "ggraph", "gganimate", "RColorBrewer",
    "ggtext", "magick", "jsonlite", "lubridate", "ggtext")

invisible(fpackage.check(packages))
rm(packages)
```

---

# import data

```{r}
data <- read.csv("./rawdata/all_apps_wide_2026-04-30-ii.csv")
times <- read.csv("./rawdata/PageTimes-2026-04-30-ii.csv")

#cbind(data$participant.node,data$participant.role)[complete.cases( #cbind(data$participant.node,data$participant.role)),]

bonus <- data[, c("participant.bonus", "participant.label")]
bonus <- bonus[!is.na(bonus$participant.bonus), ]
bonus_out <- data.frame(
  id = bonus$participant.label,
  bonus = bonus$participant.bonus
)
#nrow(bonus_out)
#write.table(bonus_out, "bonus.txt", sep = ",", row.names = FALSE, col.names = TRUE, quote = FALSE)
```

---

# results

On 30-04-2026, I recruited 59 Prolific participants via a brief [sign-up survey](https://github.com/UtrechtUniversity/sign-up-experiment) to populate a network of N=50 (with a 10% minority group). 


```{r, class.source = 'fold-hide'}
#data$session.config.network_condition
net <- jsonlite::fromJSON("./networks/network_test_n50_random.json")

g <- graph_from_adjacency_matrix(net$adj_matrix, mode = "undirected")
V(g)$role <- ifelse(net$role_vector == 1, "trendsetter", "conformist")

fplot_graph(g )
```


<br>

## diagnostics


### dropout

When did people time out?

```{r}
dropout_long <- data %>%
  filter(!is.na(participant.node)) %>%
  select(
    participant.label,
    participant.role,
    matches("unpop\\.[0-9]+\\.player\\.is_dropout")
  ) %>%
  pivot_longer(
    cols = matches("unpop\\.[0-9]+\\.player\\.is_dropout"),
    names_to = "round",
    values_to = "is_dropout"
  ) %>%
  mutate(
    round = str_extract(round, "[0-9]+"),
    round = as.numeric(round)
  )

dropout_df <- dropout_long %>%
  filter(is_dropout == 1) %>%
  select(participant.label, participant.role, round)

ggplot(na.omit(dropout_df), aes(x = round)) +
  geom_bar()

# number of players that experienced time-outs:
dropout_df %>%
  distinct(participant.label) %>%
  nrow()

# players exiting after recurring timeouts:
table(data$participant.exit)
```


----


## unpopular norm spread

```{r,class.source = 'fold-hide'}
df_long <- data %>%
  filter(!is.na(participant.node)) %>%
  select(participant.label, participant.role, participant.node, starts_with("unpop.")) %>% #also include network node.
  
  pivot_longer(
    cols = matches("unpop\\.\\d+\\.player\\.choice$"),  # only choice columns
    names_to = "round",
    values_to = "choice"
  ) %>%
  mutate(
    round = as.integer(gsub("unpop\\.(\\d+)\\.player\\.choice", "\\1", round)),
    is_bot = ifelse(participant.label == "", TRUE, FALSE)
  ) %>%
  select(participant.label, participant.node, participant.role, round, choice, is_bot)

# aggregated
df_plot <- df_long %>%
  group_by(round) %>%
  summarise(
    pct_choice1 = mean(choice, na.rm = TRUE) * 100,  # proportion * 100
    n = n()
  )

ggplot(df_plot, aes(x = round, y = pct_choice1)) +
  geom_line(group = 1, color = "steelblue", size = .5) +
  geom_point(color = "steelblue", size = 2) +
   geom_hline(yintercept = 10, linetype = "longdash", color = "darkgrey", size = 0.8) +  # dashed line at 10
  scale_x_continuous(breaks = df_plot$round) +
  scale_y_continuous(limits = c(0, 100)) +  
  labs(
    x = "Round",
    y = "% agents choosing 'blue'",
    title = "Evolution of an unpopular norm"
  )
```


```{r}
table(data$survey.1.player.enjoyment)
table(data$survey.1.player.clarity)
table(data$survey.1.player.majority)
fshowdf(unique(data$survey.1.player.strategy))
fshowdf(unique(data$survey.1.player.perceived_rq))
fshowdf(unique(data$survey.1.player.comments))
```
