---
title: "Experiment"
bibliography: references.bib
link-citations: true
date: "Last compiled on `r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    self_contained: true
    css: tweaks.css
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 4
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test3"))
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

---

# Getting started

To copy the code, click the button in the upper right corner of the code-chunks.

## clean up

```{r, clean_up, results='hide'}
rm(list=ls())
gc()
```

<br>

## custom functions

We defined a number custom functions, at `r xfun::embed_file("./custom_functions.R")`.

```{r, custom_functions}
source("./custom_functions.R")
```

<br>

## necessary packages

- `tidyverse`: data wrangling
- `igraph`: generate and visualize graphs
- `parallel`: parallel computing to speed up simulation
- `foreach`: looping in parallel
- `doParallel`: parallel backend for `foreach`
- `ggplot2`: data visualization
- `ggh4x`: hacks for `ggplot2`
- `ggpubr`: make visualizations publication-ready


```{r, packages}
packages = c("tidyverse", "igraph", "ggplot2", "parallel", "doParallel", "foreach", "ggh4x", "ggpubr", "plotly", "RColorBrewer", "grid", "gridExtra", "patchwork", "ggplotify", "ggraph", "gganimate", "RColorBrewer",
    "ggtext", "magick", "jsonlite", "lubridate", "ggtext")

invisible(fpackage.check(packages))
rm(packages)
```

---

# import data

```{r}
data <- read.csv("./rawdata/all_apps_wide_2026-03-06.csv")
times <- read.csv("./rawdata/PageTimes-2026-03-06.csv")

#cbind(data$participant.node,data$participant.role)[complete.cases( #cbind(data$participant.node,data$participant.role)),]

bonus <- data[, c("participant.bonus", "participant.label")]
bonus <- bonus[!is.na(bonus$participant.bonus), ]
bonus_out <- data.frame(
  id = bonus$participant.label,
  bonus = bonus$participant.bonus
)
nrow(bonus_out)
write.table(bonus_out, "bonus.txt", sep = ",", row.names = FALSE, col.names = TRUE, quote = FALSE)
```

---

# results

On 06-03-2026, I recruited 58 Prolific participants via a brief [sign-up survey](https://github.com/UtrechtUniversity/sign-up-experiment) to populate a network of N=50 (with a 10% minority group).

```{r, class.source = 'fold-hide'}
net <- jsonlite::fromJSON("./networks/network_test_n50.json")

g <- graph_from_adjacency_matrix(net$adj_matrix, mode = "undirected")
V(g)$role <- ifelse(net$role_vector == 1, "trendsetter", "conformist")

fplot_graph(g )
```


<br>

## diagnostics


### dropout


```{r, class.source = 'fold-hide'}
dropout_long <- data %>%
  filter(!is.na(participant.node)) %>%
  select(
    participant.label,
    participant.role,
    matches("unpop\\.[0-9]+\\.player\\.is_dropout")
  ) %>%
  pivot_longer(
    cols = matches("unpop\\.[0-9]+\\.player\\.is_dropout"),
    names_to = "round",
    values_to = "is_dropout"
  ) %>%
  mutate(
    round = str_extract(round, "[0-9]+"),
    round = as.numeric(round)
  )

dropout_df <- data.frame(
  player = 1:nrow(data),
  dropout_round = apply(
    do.call(cbind, data[paste0("unpop.",1:25,".player.is_dropout")]),
    1,
    function(x) which(x == 1)[1]
  )
)
ggplot(na.omit(dropout_df), aes(x = dropout_round)) +
  geom_bar()
```



----


## unpopular norm spread

```{r,class.source = 'fold-hide'}
df_long <- data %>%
  filter(!is.na(participant.node)) %>%
  select(participant.label, participant.role, participant.node, starts_with("unpop.")) %>% #also include network node.
  
  pivot_longer(
    cols = matches("unpop\\.\\d+\\.player\\.choice$"),  # only choice columns
    names_to = "round",
    values_to = "choice"
  ) %>%
  mutate(
    round = as.integer(gsub("unpop\\.(\\d+)\\.player\\.choice", "\\1", round)),
    is_bot = ifelse(participant.label == "", TRUE, FALSE)
  ) %>%
  select(participant.label, participant.node, participant.role, round, choice, is_bot)


# identify round of dropout:
first_dropout <- dropout_long %>%
  filter(is_dropout == 1) %>%
  group_by(participant.label) %>%
  summarise(dropout_round = min(round),
            .groups = "drop")

# and add to the df:
df_long <- df_long %>%
  left_join(first_dropout, by = "participant.label")

# aggregated
df_plot <- df_long %>%
  group_by(round) %>%
  summarise(
    pct_choice1 = mean(choice, na.rm = TRUE) * 100,  # proportion * 100
    n = n()
  )

ggplot(df_plot, aes(x = round, y = pct_choice1)) +
  geom_line(group = 1, color = "steelblue", size = .5) +
  geom_point(color = "steelblue", size = 2) +
   geom_hline(yintercept = 10, linetype = "longdash", color = "darkgrey", size = 0.8) +  # dashed line at 10
  scale_x_continuous(breaks = df_plot$round) +
  scale_y_continuous(limits = c(0, 100)) +  
  labs(
    x = "Round",
    y = "% agents choosing 'blue'",
    title = "Evolution of an unpopular norm"
  )

#sort by node
df_long <- df_long %>%
  arrange(participant.node) %>%
  select(-participant.label)

df_long$dropout_round[is.na(df_long$dropout_round)] <- 25 #non dropouts, set to 25.
```

```{r,class.source = 'fold-hide'}
# make roles consistent with utility function roles
df_long$role <- ifelse(df_long$participant.role == "Blue", "trendsetter", "conformist")

# specificy incentive structure parameters
params = list(s = 15, e = 10, w = 40, z = 50, lambda1 = 5, lambda2 = 1.8)

df_long$id <- df_long$participant.node + 1 #nodes are 0-indexed

#add degree
deg <- degree(g)
df_long <- df_long %>%
  mutate(degree = deg[id])

calculate_round_utilities <- function(current_round, df_long, network, params) {
  # previous round data
  df_prev <- df_long %>%
    filter(round == current_round - 1) %>%
    select(id, role, choice)
  
  # current round data
  df_curr <- df_long %>%
    filter(round == current_round)
  
  # compute utilities
  df_curr <- df_curr %>%
    rowwise() %>%
    mutate(
      util_0 = futility(agent_id = id, choice = 0,
                         agents = df_prev,
                         network = network,
                         params = params)$utility,
      util_1 = futility(agent_id = id, choice = 1,
                         agents = df_prev,
                         network = network,
                         params = params)$utility
    ) %>%
    ungroup()
  
  return(df_curr)
}

##calculate_round_utilities(2, df_long, network = g, params = params)

#compute utilities for all rounsd (except round 1)
max_round <- max(df_long$round)
df <- map_dfr(2:max_round, ~calculate_round_utilities(.x, df_long, g, params))

# identify best replies
df <- df %>%
  mutate(
    predicted_choice = ifelse(util_1 > util_0, 1, 0),
    best_reply = (choice == predicted_choice)  # TRUE if agent picked the choice with highest utility
  )


df_summary <- df %>%
  mutate(
    preferred_choice = ifelse(role == "trendsetter", 1, 0),    # define preferred option by role
    chose_preferred = (choice == preferred_choice)            # TRUE if they picked their preferred option
  ) %>%
  group_by(round, role) %>%
  summarize(
    n_agents = n(),
    prop_preferred = mean(chose_preferred),   # fraction that chose their preferred option
    prop_best_reply = mean(best_reply),       # fraction that picked the highest-utility choice
    .groups = "drop"
  )


ggplot(df_summary, aes(x = round)) +
  geom_line(aes(y = prop_preferred, color = "Preference"), size = 1) +
  geom_line(aes(y = prop_best_reply, color = "Best reply"), size = 1, linetype = "dashed") +
  facet_wrap(~role) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0,1)) +
  scale_color_manual(values = c("Preference" = "darkgreen", "Best reply" = "steelblue")) +
  labs(
    x = "Round",
    y = "Proportion of agents",
    color = "Metric",
    title = "Following preference vs best-reply over rounds"
  )

``` 


```{r}
table(data$survey.1.player.enjoyment)
table(data$survey.1.player.clarity)
table(data$survey.1.player.majority)
fshowdf(unique(data$survey.1.player.strategy))
fshowdf(unique(data$survey.1.player.perceived_rq))
fshowdf(unique(data$survey.1.player.comments))

#completion_code:CI5BFLAB

```
