---
title: "Experiment"
bibliography: references.bib
link-citations: true
date: "Last compiled on `r format(Sys.time(), '%d-%m-%Y')`"
output: 
  html_document:
    self_contained: true
    css: tweaks.css
    toc: true
    toc_float: true
    number_sections: true
    toc_depth: 4
    code_folding: show
    code_download: yes
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test3"))
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

---

# Getting started

To copy the code, click the button in the upper right corner of the code-chunks.

## clean up

```{r, clean_up, results='hide'}
rm(list=ls())
gc()
```

<br>

## custom functions

We defined a number custom functions, at `r xfun::embed_file("./custom_functions.R")`.

```{r, custom_functions}
source("./custom_functions.R")
```

<br>

## necessary packages

- `tidyverse`: data wrangling
- `igraph`: generate and visualize graphs
- `parallel`: parallel computing to speed up simulation
- `foreach`: looping in parallel
- `doParallel`: parallel backend for `foreach`
- `ggplot2`: data visualization
- `ggh4x`: hacks for `ggplot2`
- `ggpubr`: make visualizations publication-ready

```{r, packages}
packages = c("tidyverse", "igraph", "ggplot2", "parallel", "doParallel", "foreach", "ggh4x", "ggpubr", "plotly", "RColorBrewer", "grid", "gridExtra", "patchwork", "ggplotify", "ggraph", "gganimate", "RColorBrewer",
    "ggtext", "magick", "jsonlite")

invisible(fpackage.check(packages))
rm(packages)
```

---

# Experimental conditions


```{r, echo=TRUE, fig.show='hold', fig.keep='all', message=FALSE, fig.height=5}
# pick one configuration that likely leads to an unpopular norm, and explore multiple 'seeds':
run_one_seed <- function(
  i,
  base_seed = 1253281123,
  params = list(s = 15, e = 10, w = 40, z = 50, lambda1 = 5, lambda2 = 1.8),
  
  # tweak network
  k_min = 2, 
  k_max = 9,
  alpha = 2.1,
  rho = 0.4,
  r = -0.1,
  
  # retrieve network
  return_network = FALSE
  
) {
  # derived seed for this run
  seed_i <- base_seed + i
  set.seed(seed_i)

  # --- network creation ---
  degseq <- fdegseq(
    n      = 10,
    alpha  = alpha,
    k_min  = k_min,
    k_max  = k_max,
    dist   = "log-normal", #use log-normal 
    seed   = seed_i
  )

  network <- sample_degseq(degseq, method = "vl")
  
  V(network)$role <- sample(
    c(rep("trendsetter", 2), rep("conformist", 8))
  )
  

  rewired_network <- frewire_r(network, r, verbose = FALSE, max_iter = 1e5)
  final_network   <- fswap_rho(rewired_network, rho, verbose = FALSE, max_iter = 1e4)
  
  # --- stats ---
  stats <- list(
    run           = i,
    seed          = seed_i,
    num_nodes     = vcount(final_network),
    num_edges     = ecount(final_network),
    avg_degree    = mean(degree(final_network)),
    sd_degree     = sd(degree(final_network)),
    net_density   = edge_density(final_network),
    net_diameter  = diameter(final_network, directed = FALSE, unconnected = TRUE),
    avg_path_len  = average.path.length(final_network, directed = FALSE),
    clust_coeff   = transitivity(final_network, type = "global"),
    assort_deg    = assortativity_degree(final_network),
    deg_trait_cor = fdegtraitcor(final_network)$cor,
    components    = components(final_network)$no
  )
  
  fplot_graph(final_network, layout = layout_with_fr(final_network)) 
  
  
  # --- initial actions ---
  V(final_network)$action <- ifelse(V(final_network)$role == "trendsetter", 1, 0)
  
  # --- deterministic simulation ---
  sim_det <- fabm(
    network      = final_network,
    params       = params,
    max_rounds   = 50,
    mi_threshold = 0.49,
    choice_rule  = "deterministic",
    plot         = TRUE,
    histories    = TRUE
  )
  
  # generate the gif for the current network
  gif_filename <- paste0("./figures/animation_network_", seed_i, ".gif")
  gif_path <- fnetworkgif(final_network, sim_det$decision_history, rounds = sim_det$equilibrium$round, output_dir = "./figures")
  # rename the gif to match the naming pattern
  file.rename(gif_path, gif_filename)

  if (!is.null(sim_det$plot)) {
    print(sim_det$plot)
  }
  
  # --- probabilistic simulation ---
  sim_prob <- fabm(
    network                = final_network,
    params                 = params,
    max_rounds             = 100,
    mi_threshold           = 0.49,
    choice_rule            = "probabilistic",
    stable_window          = 8,   # the length of the window of adoption values
    required_stable_rounds = 20, # number of windows needed to declare equilibrium
    plot                   = TRUE
  )
  if (!is.null(sim_prob$plot)) {
    print(sim_prob$plot)
  }
  
   result <- list(
    segregation_det  = sim_det$equilibrium$segregation,
    segregation_prob = sim_prob$equilibrium$segregation,
    stats            = stats
  )
  
  if (return_network) {
    result$network <- final_network
  }
  
  result
}

####

run_one_sw <- function(
  i,
  base_seed = 1253281123,
  model = "watts-strogatz",
  beta = 0,
  nei = 3,
  clique_size = 5,
  pmin = 0.1,
  
  params = list(s = 15, e = 10, w = 40, z = 50, lambda1 = 5, lambda2 = 1.8),
  
  # retrieve network
  return_network = FALSE
  
) {
  # derived seed for this run
  seed_i <- base_seed + i
  set.seed(seed_i)

  if (model == "watts-strogatz") {
    network <- sample_smallworld(dim = 1, size = 10, nei = 3, p = beta) }
  else if (model == "caveman") {
    network <- simulate_caveman(n = 10, clique_size = clique_size) 
  }

  V(network)$role <- sample(
    c(rep("trendsetter", 2), rep("conformist", 8))
  )
  
  
  final_network <- network

  
  # --- stats ---
  stats <- list(
    run           = i,
    seed          = seed_i,
    num_nodes     = vcount(final_network),
    num_edges     = ecount(final_network),
    avg_degree    = mean(degree(final_network)),
    sd_degree     = sd(degree(final_network)),
    net_density   = edge_density(final_network),
    net_diameter  = diameter(final_network, directed = FALSE, unconnected = TRUE),
    avg_path_len  = average.path.length(final_network, directed = FALSE),
    clust_coeff   = transitivity(final_network, type = "global"),
    assort_deg    = assortativity_degree(final_network),
    deg_trait_cor = fdegtraitcor(final_network)$cor,
    components    = components(final_network)$no
  )
  
  stats_df <- data.frame(
    Metric = names(stats),
    Value  = unlist(stats),
    row.names = NULL
  )
  
  #print(stats_df)
  fplot_graph(final_network, layout = layout.kamada.kawai(final_network))
  

  # --- initial actions ---
  V(final_network)$action <- ifelse(V(final_network)$role == "trendsetter", 1, 0)
  
  # --- deterministic simulation ---
  sim_det <- fabm(
    network      = final_network,
    params       = params,
    max_rounds   = 50,
    mi_threshold = 0.49,
    choice_rule  = "deterministic",
    plot         = TRUE,
    histories    = TRUE
  )

  # generate the gif for the current network
  gif_filename <- paste0("./figures/animation_network_", seed_i, ".gif")
  gif_path <- fnetworkgif(final_network, sim_det$decision_history, rounds = sim_det$equilibrium$round, output_dir = "./figures")
  # rename the gif to match the naming pattern
  file.rename(gif_path, gif_filename)
  
  
  if (!is.null(sim_det$plot)) {
    print(sim_det$plot)
  }
  
  # --- probabilistic simulation ---
  sim_prob <- fabm(
    network                = final_network,
    params                 = params,
    max_rounds             = 100,
    mi_threshold           = 0.49,
    choice_rule            = "probabilistic",
    stable_window          = 8,   # the length of the window of adoption values
    required_stable_rounds = 20, # number of windows needed to declare equilibrium
    plot                   = TRUE
  )
  if (!is.null(sim_prob$plot)) {
    print(sim_prob$plot)
  }
  
  
  result <- list(
    segregation_det  = sim_det$equilibrium$segregation,
    segregation_prob = sim_prob$equilibrium$segregation,
    stats            = stats
  )
  
  if (return_network) {
    result$network <- final_network
  }
  
  result
}
```


```{r}
test <- run_one_seed(1, k_min = 2, k_max = 9, alpha = 2.1, rho = 0.7, r = -0.1, return_network = TRUE)

base = 1253281123
seed = base + 1
``` 

```{r, eval = FALSE}
# use this as the network structure for an otree session:
#cbind(degree(test$network),V(test$network)$role)

# convert to adjacency matrix
adj_matrix <- as.matrix(as_adjacency_matrix(test$network))

#get roles
role_vector <- ifelse(V(test$network)$role == "trendsetter",1,0)
# create a list to store the network data
net <- list(adj_matrix = adj_matrix, role_vector = role_vector)
# save the list as a JSON file
write_json(net, "network_test_n10.json")
```


```{r}
#random
test <- run_one_sw(i=15, beta=1, pmin=.15, return_network = TRUE )
#fplot_graph(test$network)
```
```{r,eval=FALSE}
# use this as the network structure for an otree session:
#cbind(degree(test$network),V(test$network)$role)

# convert to adjacency matrix
adj_matrix <- as.matrix(as_adjacency_matrix(test$network))

#get roles
role_vector <- ifelse(V(test$network)$role == "trendsetter",1,0)
# create a list to store the network data
net <- list(adj_matrix = adj_matrix, role_vector = role_vector)
# save the list as a JSON file
write_json(net, "network_test_n10_random.json")
```
